Researchers collaborate with same-gendered colleagues more often than expected across the life sciences

### 2018-11-01 manuscript/Manuscript_Claire.tex
Ready for resubmission

\begin{document}
\maketitle
\begin{abstract}
Evidence suggests that women in academia are hindered by conscious and
unconscious biases, and often feel excluded from formal and informal
opportunities for research collaboration. In addition to ensuring
fairness and helping to redress gender imbalance in the academic
workforce, increasing women's access to collaboration could help
scientific progress by drawing on more of the available human capital.
Here, we test whether researchers tend to collaborate with same-gendered
colleagues, using more stringent methods and a larger dataset than in
past work. Our results reaffirm that researchers co-publish with
colleagues of the same gender more often than expected by chance, and
show that this `gender homophily' is slightly stronger today than it was
10 years ago. Contrary to our expectations, we found no evidence that
homophily is driven mostly by senior academics, and no evidence that
homophily is stronger in fields where women are in the minority.
Interestingly, journals with a high impact factor for their discipline
tended to have comparatively low homophily, as predicted if mixed-gender
teams produce better research. We discuss some potential causes of
gender homophily in science. \vspace{5mm}

\par

\noindent \textbf{Keywords:} Gender bias, Homophily, Research
collaboration, Text mining, Women in STEM.
\end{abstract}

\maketitle\unmarkedfntext{*School of BioSciences, The University of Melbourne, Victoria, Australia.}
\maketitle\unmarkedfntext{\textsuperscript{§}Organismal and Evolutionary Biology Research Programme,
Faculty of Biological and Environmental Sciences, University of Helsinki, Finland.}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{gender-homophily-by-discipline-time-period-and-authorship-position}{%
\subsection{Gender homophily by discipline, time period, and authorship
position}\label{gender-homophily-by-discipline-time-period-and-authorship-position}}

\autoref{alpha_histograms} shows the distribution of \(\alpha'\)
estimates in 2015-2016 across all journals for which we recovered
sufficient data, when \(\alpha'\) was calculated for all authors, first
authors only, or last authors only. Most journals had positive values of
\(\alpha'\) (77-92\%, depending on time period and author type; S1
Data), and for many of these the false discovery rate (FDR)-corrected
p-values suggested that \(\alpha'\) was significantly greater than zero
(1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6;
S1 Data). Only 2/2077 journals had statistically significant heterophily
(i.e. \(\alpha' < 0\)) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The
remaining 606 or 787 journals (in 2015 and 2005 respectively) had a
value of \(\alpha'\) not significantly different from zero, consistent
with the null hypothesis of random assortment with respect to gender. We
also confirmed that in most journals (S2 Data) and most research
disciplines (S3 Data, S1 Fig), the majority of papers had multiple
authors. To verify the argument put forward in Figure 1, we also
calculated \(\alpha\) across all two-author papers in our entire 15-year
PubMed dataset (excluding single-author papers, and papers where at
least one author's gender was unknown; n = \textgreater{}3 million
papers, \textgreater{}16 million authors). Across the whole dataset,
\(\alpha\) was 0.126, which is almost double the median value for
individual journals (Figure 2; \(\alpha = 0.070\) for `All authors'),
suggesting that lumping together papers from different fields and
different time periods can indeed inflate \(\alpha\).

\begin{figure}
\centering
\includegraphics{../figures/Fig2.pdf}
\caption{Of the 2116 journals for which we had adequate data in
2015-2016, 825 showed statistically significant evidence of gender
homophily (denoted by \(\alpha' > 0\)), and 1 showed statistically
significant evidence of heterophily (\(\alpha' < 0\)), after false
discovery rate correction. In the stacked desnity plot, the white area
shows the number of journals for which homophily was significantly
stronger than expected under the null hypothesis (corrected p
\textless{} 0.05), while the blue area shows all the remainder. Patterns
were similar whether \(\alpha'\) was calculated for all authors, for
first authors only, or for last authors only. Points in the right panel
show \(\alpha'\) for individual journals.\label{alpha_histograms}}
\end{figure}

@@

\hypertarget{variance-in-homophily-between-disciplines}{%
\subsection{Variance in homophily between
disciplines}\label{variance-in-homophily-between-disciplines}}

@@

\hypertarget{relationship-between-gender-homophily-and-number-of-authors}{%
\subsection{Relationship between gender homophily and number of
authors}\label{relationship-between-gender-homophily-and-number-of-authors}}

Papers with two authors had significantly lower (but still positive)
\(\alpha'\) values relative to papers with more than two authors, on
average (\autoref{author_number}; statistical results in Online
Supplmentary Material). Papers with 3, 4 or 5+ authors had essentially
identical average \(\alpha'\) values. The variance in \(\alpha'\) across
journals was also a little higher for 2-authors papers compared to the
remainder (\autoref{author_number}), though part of this variance is due
to the reduced sample size (in terms of number of authors) for the
2-author papers. One possible explanation for this finding is that
2-authors papers might be more likely to have an author list that is
evenly split between career stages (e.g.~a postgraduate student and
their supervisor), increasing the chance that the authors are mixed
gender (see Figure 6). The result also suggests that the processes
responsible for gender homophily are similar in small (e.g.~3 author)
and larger (5+ authors) collaborations (and across disciplines where
small versus large collaborations are the norm).

@@

\hypertarget{relationship-between-gender-homophily-and-gender-ratio}{%
\subsection{Relationship between gender homophily and gender
ratio}\label{relationship-between-gender-homophily-and-gender-ratio}}

@@

\hypertarget{relationship-between-journal-impact-factor-and-gender-homophily}{%
\subsection{Relationship between journal impact factor and gender
homophily}\label{relationship-between-journal-impact-factor-and-gender-homophily}}

@@

\hypertarget{analysis-accounting-for-differences-in-author-gender-ratio-between-countries}{%
\subsection{Analysis accounting for differences in author gender ratio
between
countries}\label{analysis-accounting-for-differences-in-author-gender-ratio-between-countries}}

@@

\hypertarget{theoretical-expectations-for-alpha-when-the-gender-ratio-differs-between-career-stages}{%
\subsection{\texorpdfstring{Theoretical expectations for \(\alpha\) when
the gender ratio differs between career
stages}{Theoretical expectations for \textbackslash{}alpha when the gender ratio differs between career stages}}\label{theoretical-expectations-for-alpha-when-the-gender-ratio-differs-between-career-stages}}

Given that we cannot identify individual researchers or their career
stages, we used a simple model to derive the theoretical expectations
for \(\alpha\) when the gender ratio differs between career stages. As
shown in \autoref{simulation}, we predict that \(\alpha\) is expected to
be non-zero, even if collaborators are randomly selected with respect to
gender, provided that there is a gender gap between career stages. The
extent to which \(\alpha\) deviates from zero depends on the relative
frequencies of collaboration within and between career stages (rows and
columns in Figure 6), and the size of the gender gap between stages (x-
and y-axes in Figure 6). When \textgreater{}50\% of coauthor pairs
comprise one early-career and one established researcher, we expect
gender heterophily (\(\alpha < 0\)) whenever the gender ratio differs
between career stages. Conversely, when \textgreater{}50\% of
collaborations are between people at the same career stage, we expect
gender homophily (\(\alpha > 0\)). In a few parameter spaces (shown in
red; \autoref{simulation}), \(\alpha\) was quite high, and overlapped
with the values that we estimated (\autoref{alpha_histograms}).

\begin{figure}
\centering
\includegraphics{../figures/Fig6_inkscape.pdf}
\caption{When there is a difference in gender ratio between early-career
and established researchers, and collaboration is non-random with
respect to career stage, the null expectation for \(\alpha\) deviates
from zero. An excess of collaborations between career stages gives the
appearance of gender heterophily (lower rows, blue areas), while an
excess of within-career stage collaborations produced apparent gender
homophily (upper rows, red areas). However, the conditions required for
strong gender homophily are quite restrictive, making it unlikely that
this issue explains all of the homophily observed in Figure 2. Contour
lines denote increments of 0.1. \label{simulation}}
\end{figure}

@@

Lastly, we note that if there is a gender gap between career stages and
coauthorships between early-career and established researchers comprise
\textgreater{}50\% of the total, then the baseline expectation for
\(\alpha\) is actually less than zero (blue areas in
\autoref{simulation}). Therefore, it is possible that researchers
preferentially assort with same-gendered collaborators even more
strongly than implied by other results, at least for certain journals or
research disciplines.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@

Regarding our finding of weaker homophily among 2-author papers, we
suspect that many 2-author teams comprise a student/postdoc and a senior
staff member, making these teams especially likely to be mixed-gender,
due to the greater shortage of women among senior researchers {[}1,5{]}.
Assuming this interpretation is correct, this result suggests that our
reported \(\alpha'\) values may underestimate the strength of peoples'
preferences for same-gendered collaborators; essentially, women seeking
a senior collaborator could be constrained to work mostly with men,
meaning that people's ideal and realised gender preferences would be
mismatched. On a related note, Ghiasi et al. {[}51{]} argue that women
in engineering are ``compliant {[}in reproducing{]} male-dominated
scientific structures'' because they do not collaborate often enough
with other women (for reference, their Figure 7 implies that
coauthorships involving two women are 30\% more frequent than expected
under random assortment). By contrast, we feel that it may be
counter-productive to recommend that women collaborate primarily with
other women, e.g.~because this constrains women's options (particularly
in fields like engineering, where 90\% of professors are men {[}1{]}).
Instead, we suggest that researchers of both genders can help to close
the gender gap in STEMM. In the context of collaboration, one way to do
this is to undertake self-examination to ensure that one is not
inadvertently overlooking or excluding women among potential students
and colleagues. One should also take care to treat male and female
collaborators equally, e.g.~in terms of training and mentoring,
allocation of work, how one frames or promotes the collaboration
(e.g.~in conference presentations or on a website), and in terms of the
credit afforded to collaborators' contributions. Experimental work
suggests that unconscious bias causes people to undervalue women's
research achievements {[}20{]}, and a study of author contribution
statements found observational evidence that menial or under-valued
tasks are assigned to women while more prestigious tasks are assigned to
men {[}61{]}.

Our study begs two questions: what causes gender homophily in science,
and are our results cause for concern? These questions are closely
related. For example, some of the homophily we observed might be caused
by women seeking to avoid harassment or sexism from men {[}38{]}, which
would clearly be concerning. Additionally, Sheltzer and Smith {[}66{]}
concluded that `elite' male academics (defined as recipients of major
honours) have a higher proportion of male students and postdocs than
non-elite male academics. This finding could contribute to the homophily
we observed, and is cause for concern since the results might reflect
discrimination against women during hiring {[}20{]}, or avoidance by
women of elite research groups (e.g.~due to gender differences in
confidence, or a perception that some groups are sexist). We also found
a little evidence that gender homophily is detrimental to research
quality, in that high-impact journals tended to have weaker homophily.
Assuming that papers published in high-impact journals are of higher
average quality (which is contentious; {[}67{]}), our results provide
non-experimental support for the hypothesis that mixed-gender teams
produce better research than single-gender teams {[}42--48{]}. Another
issue is that if many collaborations are between established
researchers, there will be an excess of male-male collaborations in
fields where women in senior positions are rare; some of the observed
homophily might therefore reflect the elevated gender gap among senior
researchers.

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{the-dataset}{%
\subsection{The dataset}\label{the-dataset}}

@@

\hypertarget{calculating-alpha-the-coefficient-of-homophily}{%
\subsection{\texorpdfstring{Calculating \(\alpha\), the coefficient of
homophily}{Calculating \textbackslash{}alpha, the coefficient of homophily}}\label{calculating-alpha-the-coefficient-of-homophily}}

@@

\hypertarget{minimising-the-wahlund-effect-research-discipline-and-time-period}{%
\subsection{Minimising the Wahlund effect: research discipline and time
period}\label{minimising-the-wahlund-effect-research-discipline-and-time-period}}

@@

\hypertarget{minimising-the-wahlund-effect-author-country-of-affiliation}{%
\subsection{Minimising the Wahlund effect: author country of
affiliation}\label{minimising-the-wahlund-effect-author-country-of-affiliation}}

@@

\hypertarget{calculating-standardised-journal-impact-factor}{%
\subsection{Calculating standardised journal impact
factor}\label{calculating-standardised-journal-impact-factor}}

@@

\hypertarget{statistical-analysis}{%
\subsection{Statistical analysis}\label{statistical-analysis}}

@@

\hypertarget{theoretical-expectations-for-alpha-when-the-gender-ratio-differs-between-career-stages-1}{%
\subsection{\texorpdfstring{Theoretical expectations for \(\alpha\) when
the gender ratio differs between career
stages}{Theoretical expectations for \textbackslash{}alpha when the gender ratio differs between career stages}}\label{theoretical-expectations-for-alpha-when-the-gender-ratio-differs-between-career-stages-1}}

In most STEMM subjects, the gender ratio is more skewed among
established researchers relative to early-career researchers, due both
to women leaving STEMM careers at greater rates (the `leaky pipeline'),
and to historical shortages of women studying STEMM (`demographic
inertia') {[}1,5{]}. We hypothesised that this difference in gender
ratio between career stages could potentially create both Wahlund
effects and `reverse' Wahlund effects. For example, imagine that the
majority of collaborations in a particular field are between students
and professors, and that the gender ratio differs between career stages:
we would then see an excess of mixed-gender coauthorships (heterophily,
\(\alpha < 0\)), even if gender has no direct, causal effect. Similarly,
a hypothetical field in which students work only with students, and
professors with professors, would have apparent gender homophily
(\(\alpha > 0\)).

@@

\hypertarget{data-availability-and-reproducibility}{%
\subsection{Data availability and
reproducibility}\label{data-availability-and-reproducibility}}

@@

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

@@

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Shaw_2012}{}%
1. Shaw AK, Stanton DE. Leaks in the pipeline: separating demographic
inertia from ongoing gender differences in academia. Proceedings of the
Royal Society of London B. 2012;272: 3736--3741.

\leavevmode\hypertarget{ref-Lariviere_2013}{}%
2. Larivière V, Ni C, Gingras Y, Cronin B, Sugimoto CR. Bibliometrics:
global gender disparities in science. Nature. 2013;504: 211--213.

\leavevmode\hypertarget{ref-West_2013}{}%
3. West JD, Jacquet J, King MM, Correll SJ, Bergstrom CT. The role of
gender in scholarly authorship. PLoS ONE. 2013;8: e66212.

\leavevmode\hypertarget{ref-Elsevier_report}{}%
4. Elsevier Report. Gender in the global research landscape.
elseviercom/research-intelligence/resource-library/gender-report. 2017;

\leavevmode\hypertarget{ref-Holman_2018}{}%
5. Holman L, Stuart Fox D, Hauser CE. The gender gap in science: How
long until women are equally represented? PLoS Biology. 2018;16:
e2004956.

\leavevmode\hypertarget{ref-Wutte_2007}{}%
6. Wutte M. Closing the gender gap. Nature. 2007;448: NJ101--NJ102.

\leavevmode\hypertarget{ref-Reuben_2014}{}%
7. Reuben E, Sapienza P, Zingales L. How stereotypes impair women's
careers in science. Proceedings of the National Academy of Sciences.
2014;111: 4403--4408.

\leavevmode\hypertarget{ref-Trower_2002}{}%
8. Trower CA, Chait RP. Faculty diversity: Why women and minorities are
underrepresented in the professoriate, and fresh ideas to induce needed
reform. Harvard Magazine. 2002;104: 33--37.

\leavevmode\hypertarget{ref-Umbach_2007}{}%
9. Umbach PD. Gender equity in the academic labor market: An analysis of
academic disciplines. Research in Higher Education. 2007;48: 169--192.

\leavevmode\hypertarget{ref-Hosek_2005}{}%
10. Hosek S, Cox AG, Ghosh-Dastidar B, Kofner A, Ramphal N, Scott J, et
al. Gender differences in major federal external grant programs. RAND
Corporation. 2005;

\leavevmode\hypertarget{ref-pohlhaus_2011}{}%
11. Pohlhaus JR, Jiang H, Wagner RM, Schaffer WT, Pinn VW. Sex
differences in application, success, and funding rates for NIH
extramural programs. Academic Medicine. 2011;86: 759.

\leavevmode\hypertarget{ref-Zuckerman_1987}{}%
12. Zuckerman H. Persistence and change in the careers of men and women
scientists and engineers. National Academy Press. 1987; 127--156.

\leavevmode\hypertarget{ref-Rosenfeld_1991}{}%
13. Rosenfeld RA. Outcome analysis of academic careers. Review prepared
for the Office of Scientific and Engineering Personnel, National
Research Council. 1991;

\leavevmode\hypertarget{ref-Long_1993}{}%
14. Long JS, Paul DA, Robert M. Rank advancement in academic careers:
Sex differences and the effects of productivity. American Sociological
Review. 1993; 703--722.

\leavevmode\hypertarget{ref-Hopkins_2013}{}%
15. Hopkins AL, Jawitz JW, McCarty C, Goldman A, Basu NB. Disparities in
publication patterns by gender, race and ethnicity based on a survey of
a random sample of authors. Scientometrics. 2013;96: 515--534.

\leavevmode\hypertarget{ref-ODorchai_2009}{}%
16. O'Dorchai S, Meulders D, Crippa F, Margherita A. She figures
2009--Statistics and indicators on gender equality in science.
Publications Office of the European Union. 2009;

\leavevmode\hypertarget{ref-Feldt_1986}{}%
17. Feldt B. The faculty cohort study: School of medicine. Ann Arbor,
MI: Office of Affirmative Action. 1986;

\leavevmode\hypertarget{ref-Stack_2004}{}%
18. Stack S. Gender, children and research productivity. Scientometrics.
2004;45: 891--920.

\leavevmode\hypertarget{ref-Lariviere_2011}{}%
19. Larivière V, Vignola-Gagné E, Villeneuve C, Gélinas P, Gingras Y.
Sex differences in research funding, productivity and impact: an
analysis of Québec university professors. Scientometrics. 2011;87:
483--498.

\leavevmode\hypertarget{ref-Moss_2012}{}%
20. Moss-Racusin CA, Dovidio JF, Brescoll VL, Graham MJ, Handelsman J.
Science faculty's subtle gender biases favor male students. Proceedings
of the National Academy of Sciences. 2012;109: 16474--16479.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Evidence suggests that women in academia are hindered by conscious and
unconscious biases, and often feel excluded from formal and informal
opportunities for research collaboration. In addition to ensuring
fairness and helping to redress gender imbalance in the academic
workforce, increasing women's access to collaboration could help
scientific progress by drawing on more of the available human capital.
Here, we test whether researchers tend to collaborate with same-gendered
colleagues, using more stringent methods and a larger dataset than in
past work. Our results reaffirm that researchers co-publish with
colleagues of the same gender more often than expected by chance, and
show that this `gender homophily' is slightly stronger today than it was
10 years ago. Contrary to our expectations, we found no evidence that
homophily is driven mostly by senior academics, and no evidence that
homophily is stronger in fields where women are in the minority.
Interestingly, journals with a high impact factor for their discipline
tended to have comparatively low homophily, as predicted if mixed-gender
teams produce better research. We discuss some potential causes of
gender homophily in science. \vspace{5mm}
\par\noindent \textbf{Keywords:} Gender bias, Homophily, Research
collaboration, Text mining, Women in STEM.
\end{abstract}

\maketitle

\unmarkedfntext{*School of BioSciences, The University of Melbourne, Victoria, Australia.}
\maketitle\unmarkedfntext{\textsuperscript{§}Organismal and Evolutionary Biology Research Programme,
Faculty of Biological and Environmental Sciences, University of Helsinki, Finland.}

@@

\section{Introduction}\label{introduction}

@@

\section{Results}\label{results}

\subsection{Gender homophily by discipline, time period, and authorship
position}\label{gender-homophily-by-discipline-time-period-and-authorship-position}

\autoref{alpha_histograms} shows the distribution of \(\alpha'\)
estimates in 2015-2016 across all journals for which we recovered
sufficient data, when \(\alpha'\) was calculated for all authors, first
authors only, or last authors only. Most journals had positive values of
\(\alpha'\) (77-92\%, depending on time period and author type; S1
Data), and for many of these the false discovery rate (FDR)-corrected
p-values suggested that \(\alpha'\) was significantly greater than zero
(1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6;
S1 Data). Only 2/2077 journals had statistically significant heterophily
(i.e. \(\alpha' < 0\)) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The
remaining 606 or 787 journals (in 2015 and 2005 respectively) had a
value of \(\alpha'\) not significantly different from zero, consistent
with the null hypothesis of random assortment with respect to gender. We
also confirmed that in most journals (S2 Data) and most research
disciplines (S3 Data, S1 Fig), the majority of papers had multiple
authors.

\begin{figure}
\centering
\includegraphics{../figures/Fig2.pdf}
\caption{Of the 2116 journals for which we had adequate data in
2015-2016, 825 showed statistically significant evidence of gender
homophily (denoted by \(\alpha' > 0\)), and 1 showed statistically
significant evidence of heterophily (\(\alpha' < 0\)), after false
discovery rate correction. In the stacked density plot, the white area
shows the number of journals for which homophily was significantly
stronger than expected under the null hypothesis (corrected p
\textless{} 0.05), while the blue area shows all the remainder. Patterns
were similar whether \(\alpha'\) was calculated for all authors, for
first authors only, or for last authors only. Points in the right panel
show \(\alpha'\) for individual journals.\label{alpha_histograms}}
\end{figure}

@@

\subsection{Variance in homophily between
disciplines}\label{variance-in-homophily-between-disciplines}

@@

Nevertheless, when we calculated \(\alpha\) across all non-single-author
papers in our entire 15-year PubMed dataset (as before, excluding papers
where at least one author's gender was unknown; n = \textgreater{}3
million papers, \textgreater{}16 million authors), we found that
\(\alpha\) was 0.126. This figure is almost double the median value of
\(\alpha'\) for individual journals (Figure 2; \(\alpha' = 0.070\) for
`All authors'), suggesting that lumping together papers from different
fields and different time periods can indeed produce spurious evidence
for gender homophily as outlined in Figure 1.

\subsection{Relationship between gender homophily and number of
authors}\label{relationship-between-gender-homophily-and-number-of-authors}

Papers with two authors had significantly lower (but still positive)
\(\alpha'\) values relative to papers with more than two authors, on
average (\autoref{author_number}; statistical results in Online
Supplmentary Material). Papers with 3, 4 or \(\ge5\) authors had
essentially identical average \(\alpha'\) values. The variance in
\(\alpha'\) across journals was also a little higher for 2-authors
papers compared to the remainder (\autoref{author_number}), though part
of this variance is due to the reduced sample size (in terms of number
of authors) for the 2-author papers. One possible explanation for this
finding is that 2-authors papers might be more likely to have an author
list that is evenly split between career stages (e.g.~a postgraduate
student and their supervisor), increasing the chance that the authors
are mixed gender (see Figure 6). The result also suggests that the
processes responsible for gender homophily are similar in small (e.g.~3
author) and larger (5+ authors) collaborations (and across disciplines
where small versus large collaborations are the norm).

@@

\subsection{Relationship between gender homophily and gender
ratio}\label{relationship-between-gender-homophily-and-gender-ratio}

@@

\subsection{Relationship between journal impact factor and gender
homophily}\label{relationship-between-journal-impact-factor-and-gender-homophily}

@@

\subsection{Analysis accounting for differences in author gender ratio
between
countries}\label{analysis-accounting-for-differences-in-author-gender-ratio-between-countries}

@@

\subsection{\texorpdfstring{Theoretical expectations for \(\alpha\) when
the gender ratio differs between career
stages}{Theoretical expectations for \textbackslash{}alpha when the gender ratio differs between career stages}}\label{theoretical-expectations-for-alpha-when-the-gender-ratio-differs-between-career-stages}

Given that we cannot identify individual researchers or their career
stages, we used a simple model to derive the theoretical expectations
for \(\alpha\) when the gender ratio differs between career stages (see
Methods). As shown in \autoref{simulation}, we predict that \(\alpha\)
is expected to be non-zero, even if collaborators are randomly selected
with respect to gender, provided that there is a gender gap between
career stages. The extent to which \(\alpha\) deviates from zero depends
on the relative frequencies of collaboration within and between career
stages (rows and columns in Figure 6), and the size of the gender gap
between stages (x- and y-axes in Figure 6). When \textgreater{}50\% of
coauthor pairs comprise one early-career and one established researcher,
we expect gender heterophily (\(\alpha < 0\)) whenever the gender ratio
differs between career stages. Conversely, when \textgreater{}50\% of
collaborations are between people at the same career stage, we expect
gender homophily (\(\alpha > 0\)). In a few parameter spaces (shown in
red; \autoref{simulation}), \(\alpha\) was quite high, and overlapped
with the values that we estimated (\autoref{alpha_histograms}).

\begin{figure}
\centering
\includegraphics{../figures/Fig6_inkscape.pdf}
\caption{When the gender ratio of early-career researchers is not equal
to the gender ratio among established researchers, the null expectation
for \(\alpha\) is not necessarily zero. Specifically, if most
collaborations occur between career stages, there will be an excess of
mixed-gender collaborations (\(\alpha < 0\), blue areas), while if most
collaborator pairs comprise two people at the same career stage, there
will be an excess of same-gender collaborations (\(\alpha > 0\), red
areas). However, the conditions required for strong gender homophily
(i.e.~the red areas) are quite restrictive, making it unlikely that this
issue can fully explain the homophily observed in our study.
Additionally, in research disciplines where between-career stage
collaboration is common and there is a shortage of women among
established researchers (i.e.~the blue areas), our study will
underestimate the strength of gender homophily. Contour lines denote
increments of 0.1. \label{simulation}}
\end{figure}

@@

Lastly, we note that if there is a gender gap between career stages and
coauthorships between early-career and established researchers comprise
\textgreater{}50\% of the total, then the baseline expectation for
\(\alpha\) is actually less than zero (blue areas in
\autoref{simulation}). Therefore, it is possible that researchers
preferentially assort with same-gendered collaborators even more
strongly than implied by our results, at least for certain journals or
research disciplines.

\section{Discussion}\label{discussion}

@@

Regarding our finding of weaker homophily among 2-author papers, we
suspect that many 2-author teams comprise a student/postdoc and a senior
staff member, making these teams especially likely to be mixed-gender,
due to the greater shortage of women among senior researchers {[}1,5{]}.
Assuming this interpretation is correct, this result suggests that our
reported \(\alpha'\) values may underestimate the strength of peoples'
preferences for same-gendered collaborators; essentially, women seeking
a senior collaborator could be constrained to work mostly with men,
meaning that people's ideal and realised gender preferences would be
mismatched. On a related note, Ghiasi et al. {[}51{]} argue that women
in engineering are ``compliant {[}in reproducing{]} male-dominated
scientific structures'' because they do not collaborate often enough
with other women (for reference, Figure 7 in {[}51{]} implies that
coauthorships involving two women are \(c\). 30\% more frequent than
expected under random assortment). By contrast, we feel that it may be
counter-productive to recommend that women collaborate primarily with
other women, e.g.~because this constrains women's options (particularly
in fields like engineering, where 90\% of professors are men {[}1{]}).
Instead, we suggest that researchers of both genders can help to close
the gender gap in STEMM. In the context of collaboration, one way to do
this is to undertake self-examination to ensure that one is not
inadvertently overlooking or excluding women among potential students
and colleagues. One should also take care to treat male and female
collaborators equally, e.g.~in terms of training and mentoring,
allocation of work, how one frames or promotes the collaboration
(e.g.~in conference presentations or on a website), and in terms of the
credit afforded to collaborators' contributions. Experimental work
suggests that unconscious bias causes people to undervalue women's
research achievements {[}20{]}, and a study of author contribution
statements found observational evidence that menial or under-valued
tasks are assigned to women while more prestigious tasks are assigned to
men {[}61{]}.

Our study begs two questions: what causes gender homophily in science,
and are our results cause for concern? We beleive that the answers to
these questions are closely related. For example, some of the homophily
we observed might be caused by women seeking to avoid harassment or
sexism from men {[}38{]}, which would clearly be concerning.
Additionally, Sheltzer and Smith {[}66{]} concluded that `elite' male
academics (defined as recipients of major honours) have a higher
proportion of male students and postdocs than non-elite male academics.
This finding could contribute to the homophily we observed, and is cause
for concern since the results might reflect discrimination against women
during hiring {[}20{]}, or avoidance by women of elite research groups
(e.g.~due to gender differences in confidence, or a perception that some
groups are sexist). We also found a little evidence that gender
homophily is detrimental to research quality, in that high-impact
journals tended to have weaker homophily. Assuming that papers published
in high-impact journals are of higher average quality (which is
contentious; {[}67{]}), our results provide non-experimental support for
the hypothesis that mixed-gender teams produce better research than
single-gender teams {[}42--48{]}. Another issue is that if many
collaborations are between established researchers, there will be an
excess of male-male collaborations in fields where women in senior
positions are rare; some of the observed homophily might therefore
reflect the elevated gender gap among senior researchers.

@@

\section{Methods}\label{methods}

\subsection{The dataset}\label{the-dataset}

@@

\subsection{\texorpdfstring{Calculating \(\alpha\), the coefficient of
homophily}{Calculating \textbackslash{}alpha, the coefficient of homophily}}\label{calculating-alpha-the-coefficient-of-homophily}

@@

\subsection{Minimising the Wahlund effect: research discipline and time
period}\label{minimising-the-wahlund-effect-research-discipline-and-time-period}

@@

\subsection{Minimising the Wahlund effect: author country of
affiliation}\label{minimising-the-wahlund-effect-author-country-of-affiliation}

@@

\subsection{Calculating standardised journal impact
factor}\label{calculating-standardised-journal-impact-factor}

@@

\subsection{Statistical analysis}\label{statistical-analysis}

@@

\subsection{\texorpdfstring{Theoretical expectations for \(\alpha\) when
the gender ratio differs between career
stages}{Theoretical expectations for \textbackslash{}alpha when the gender ratio differs between career stages}}\label{theoretical-expectations-for-alpha-when-the-gender-ratio-differs-between-career-stages-1}

In most STEMM disciplines, the gender ratio is more skewed among
established researchers relative to early-career researchers, due both
to women leaving STEMM careers at greater rates (the `leaky pipeline'),
and to historical shortages of women studying STEMM subjects at
university (`demographic inertia') {[}1,5{]}. We hypothesised that this
difference in gender ratio between career stages could potentially
create both Wahlund effects and `reverse' Wahlund effects. For example,
imagine that the majority of collaborations in a particular field are
between students and professors, and that the gender ratio differs
between career stages: we would then see an excess of mixed-gender
coauthorships (heterophily, \(\alpha < 0\)), even if gender has no
direct, causal effect. Similarly, a hypothetical field in which students
work only with students, and professors with professors, would have
apparent gender homophily (\(\alpha > 0\)).

@@

\subsection{Data availability and
reproducibility}\label{data-availability-and-reproducibility}

@@

\section{Acknowledgements}\label{acknowledgements}

@@

\section{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-Shaw_2012}{}
1. Shaw AK, Stanton DE. Leaks in the pipeline: separating demographic
inertia from ongoing gender differences in academia. Proceedings of the
Royal Society of London B. 2012;272: 3736--3741.

\hypertarget{ref-Lariviere_2013}{}
2. Larivière V, Ni C, Gingras Y, Cronin B, Sugimoto CR. Bibliometrics:
global gender disparities in science. Nature. 2013;504: 211--213.

\hypertarget{ref-West_2013}{}
3. West JD, Jacquet J, King MM, Correll SJ, Bergstrom CT. The role of
gender in scholarly authorship. PLoS ONE. 2013;8: e66212.

\hypertarget{ref-Elsevier_report}{}
4. Elsevier Report. Gender in the global research landscape.
elseviercom/research-intelligence/resource-library/gender-report. 2017;

\hypertarget{ref-Holman_2018}{}
5. Holman L, Stuart Fox D, Hauser CE. The gender gap in science: How
long until women are equally represented? PLoS Biology. 2018;16:
e2004956.

\hypertarget{ref-Wutte_2007}{}
6. Wutte M. Closing the gender gap. Nature. 2007;448: NJ101--NJ102.

\hypertarget{ref-Reuben_2014}{}
7. Reuben E, Sapienza P, Zingales L. How stereotypes impair women's
careers in science. Proceedings of the National Academy of Sciences.
2014;111: 4403--4408.

\hypertarget{ref-Trower_2002}{}
8. Trower CA, Chait RP. Faculty diversity: Why women and minorities are
underrepresented in the professoriate, and fresh ideas to induce needed
reform. Harvard Magazine. 2002;104: 33--37.

\hypertarget{ref-Umbach_2007}{}
9. Umbach PD. Gender equity in the academic labor market: An analysis of
academic disciplines. Research in Higher Education. 2007;48: 169--192.

\hypertarget{ref-Hosek_2005}{}
10. Hosek S, Cox AG, Ghosh-Dastidar B, Kofner A, Ramphal N, Scott J, et
al. Gender differences in major federal external grant programs. RAND
Corporation. 2005;

\hypertarget{ref-pohlhaus_2011}{}
11. Pohlhaus JR, Jiang H, Wagner RM, Schaffer WT, Pinn VW. Sex
differences in application, success, and funding rates for NIH
extramural programs. Academic Medicine. 2011;86: 759.

\hypertarget{ref-Zuckerman_1987}{}
12. Zuckerman H. Persistence and change in the careers of men and women
scientists and engineers. National Academy Press. 1987; 127--156.

\hypertarget{ref-Rosenfeld_1991}{}
13. Rosenfeld RA. Outcome analysis of academic careers. Review prepared
for the Office of Scientific and Engineering Personnel, National
Research Council. 1991;

\hypertarget{ref-Long_1993}{}
14. Long JS, Paul DA, Robert M. Rank advancement in academic careers:
Sex differences and the effects of productivity. American Sociological
Review. 1993; 703--722.

\hypertarget{ref-Hopkins_2013}{}
15. Hopkins AL, Jawitz JW, McCarty C, Goldman A, Basu NB. Disparities in
publication patterns by gender, race and ethnicity based on a survey of
a random sample of authors. Scientometrics. 2013;96: 515--534.

\hypertarget{ref-ODorchai_2009}{}
16. O'Dorchai S, Meulders D, Crippa F, Margherita A. She figures
2009--Statistics and indicators on gender equality in science.
Publications Office of the European Union. 2009;

\hypertarget{ref-Feldt_1986}{}
17. Feldt B. The faculty cohort study: School of medicine. Ann Arbor,
MI: Office of Affirmative Action. 1986;

\hypertarget{ref-Stack_2004}{}
18. Stack S. Gender, children and research productivity. Scientometrics.
2004;45: 891--920.

\hypertarget{ref-Lariviere_2011}{}
19. Larivière V, Vignola-Gagné E, Villeneuve C, Gélinas P, Gingras Y.
Sex differences in research funding, productivity and impact: an
analysis of Québec university professors. Scientometrics. 2011;87:
483--498.

\hypertarget{ref-Moss_2012}{}
20. Moss-Racusin CA, Dovidio JF, Brescoll VL, Graham MJ, Handelsman J.
Science faculty's subtle gender biases favor male students. Proceedings
of the National Academy of Sciences. 2012;109: 16474--16479.

\hypertarget{ref-Knobloch_2013}{}
21. Knobloch-Westerwick S, Glynn CJ, Huge M. Science faculty's subtle
gender biases favor male students. Science Communication. 2013;35:
603--625.

\hypertarget{ref-Lee_2005}{}
22. Lee S, Bozeman B. The impact of research collaboration on scientific
productivity. Social Studies of Science. 2005;35: 673--702.
